---
title: "Heath and Economic Effects of Severe Weather"
author: "Aditya Taunk"
date: '2017-06-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Synopsis


## Data Processing

Download data if it's not found in working directory
```{r}
if(!file.exists("StormData.csv.bz2")){
  download.file(
    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
    "StormData.csv.bz2"
  )
}
```

Read it into memory
```{r}
library(data.table)
data <- fread("bzcat StormData.csv.bz2")
```

We are interested in the data broken down by event types
```{r}
evtypes <- data[ , 
  list(
    frequency = .N,
    injuries = sum(INJURIES), 
    fatalities = sum(FATALITIES),
    property.damage = sum(PROPDMG),
    crop.damage = sum(CROPDMG)
  ),
  by = EVTYPE
]
```

### Harm

Create a column which indicates how much harm a particular event type has caused.
Here harm is defined as the sum of the fractions of injuries, fatalities, property 
and crop damage the event type has caused.
```{r}
totals <- evtypes[,
  list(
    injuries = sum(injuries),
    fatalities = sum(fatalities),
    property.damage = sum(property.damage),
    crop.damage = sum(crop.damage)
  )
]
evtypes[,
  harm := 
    injuries/totals$injuries +
    fatalities/totals$fatalities +
    property.damage/totals$property.damage +
    crop.damage/totals$crop.damage
]
```

Sort event types by amount of harm and rank them by the type of harm
```{r}
evtypes <-evtypes[order(-harm)]
evranks <- evtypes[,
  list(
    EVTYPE,
    injuries = rank(-injuries, ties.method="first"),
    fatalities = rank(-fatalities, ties.method="first"),
    property.damage = rank(-property.damage, ties.method="first"),
    crop.damage = rank(-crop.damage, ties.method="first")
  )
]
```

Lets see the top 21 most harmful events. Why 21? because they contain the top 10
causes for each of the harm types.
```{r}
head(evranks, 21)
```

As seen from the above table, tornados are the leading cause of injuries, fatalities 
and property damage. They are also a significant cause of crop damage. Also interesting
to note is that event types like heat seem hazardous to health but not so much for
property and crops. Conversely, things like hail and thunderstorm winds affect crop
and property but they affect heath to a lesser extent.

The harm value is a made up value to find the most relavant EVTYPEs. We don't need 
it anymore.
```{r}
evtypes <- evtypes[, !"harm"]
```

We want to know what percent of the total harm do these 21 account for
```{r}
toptypes <- head(evtypes, 21)
toptypes[,
  list(
    injuries = sum(injuries)/totals$injuries,
    fatalities = sum(fatalities)/totals$fatalities,
    property.damage = sum(property.damage)/totals$property.damage,
    crop.damage = sum(crop.damage)/totals$crop.damage
  )
]
```

So our subset accounts for a majority of the harm reported in the data (at least 
86%). So now we will mostly focus on those 21 EVTYPEs. We will add a row to the 
subset summing up the contributions from the other EVTYPEs.
```{r}
others <- evtypes[!EVTYPE %in% toptypes$EVTYPE,]
otherstotal <- others[,
  list(
    EVTYPE = "OTHER",
    frequency = sum(frequency),
    injuries = sum(injuries),
    fatalities = sum(fatalities),
    property.damage = sum(property.damage),
    crop.damage = sum(crop.damage)
  )
]
evtypes <- rbind(toptypes,otherstotal)
options(width = 120)
evtypes
```

Here is a plot of the above data separated by the type of harm
```{r}
par(mar=c(2,3,2,1))
layout(matrix(c(1,2,5,3,4,5),2,3,byrow=T))
library(colorRamps)
cols <- primary.colors(22)

with(evtypes, barplot(injuries+1, main="Injuries", log="y", yaxt="n", col=cols))
axis(2, at=c(6,51,501,5001,50001), label=c(5,50,500,5000,50000))

with(evtypes, barplot(fatalities+1, main="Fatalities", log="y", yaxt="n", col=cols))
axis(2, at=c(1,6,51,501,5001), label=c(0,5,50,500,5000))

with(evtypes, barplot(property.damage+1, main="Property Damage ($)", yaxt="n", log="y", col=cols))
axis(2, at=c(6,51,501,10001,1000001), label=c(5,50,500,1e4,1000000))

with(evtypes, barplot(crop.damage+1, main="Crop Damage ($)", log="y", yaxt="n", col=cols))
axis(2, at=c(1,6,51,501,10001,500001), label=c(0,5,50,500,1e4,500000))

plot(1, 1, type="n", ann=FALSE, yaxt='n', xaxt='n')
legend("center", evtypes$EVTYPE, fill=cols, box.lwd=0)
```






## Results





