---
title: "Heath and Economic Effects of Severe Weather"
author: "Aditya Taunk"
date: '2017-06-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

## Synopsis

In this report we will look at what types of severe weather cause the most harm
to health (injuries and fatalities) and which types have the most economic impact
(property and crop damage). We will see how these have evolved over time. Then
we will look at Tornadoes which are a major contributor to injuries and Hail which
is a major contributor to crop damage.

## Data Processing

Download data if it's not found in working directory
```{r}
if(!file.exists("StormData.csv.bz2")){
  download.file(
    "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
    "StormData.csv.bz2"
  )
}
```

Read it into memory
```{r}
library(data.table)
data <- fread(
  "bzcat StormData.csv.bz2",
  na.strings = ""
)
```

We are only interested in the data after 1996. Before then, the only information 
we have is on tornadoes, wind and hail but we want to look at more EVTYPEs
```{r}
dates <-  as.Date(data$BGN_DATE, format="%m/%d/%Y")
data[, year := as.integer(format(dates, format="%Y"))]
data <- data[year >= 1996]
```


## Results

### Event Types

We are interested in the data broken down by event types
```{r}
evtypes <- data[ , 
  list(
    frequency = .N,
    injuries = sum(INJURIES), 
    fatalities = sum(FATALITIES),
    property.damage = sum(PROPDMG),
    crop.damage = sum(CROPDMG)
  ),
  by = EVTYPE
]
```

Create a column which indicates how much harm a particular event type has caused.
Here harm is defined as the sum of the fractions of injuries, fatalities, property 
and crop damage the event type has caused.
```{r}
totals <- evtypes[,
  list(
    injuries = sum(injuries),
    fatalities = sum(fatalities),
    property.damage = sum(property.damage),
    crop.damage = sum(crop.damage)
  )
]
evtypes[,
  harm := 
    injuries/totals$injuries +
    fatalities/totals$fatalities +
    property.damage/totals$property.damage +
    crop.damage/totals$crop.damage
]
```

Sort event types by amount of harm and rank them by the type of harm
```{r}
evtypes <-evtypes[order(-harm)]
evranks <- evtypes[,
  list(
    EVTYPE,
    injuries = rank(-injuries, ties.method="first"),
    fatalities = rank(-fatalities, ties.method="first"),
    property.damage = rank(-property.damage, ties.method="first"),
    crop.damage = rank(-crop.damage, ties.method="first")
  )
]
```

Lets see the top 10 most harmful events.
```{r}
head(evranks, 10)
```

As seen from the above table, tornados are the leading cause of injuries and a 
significant cause of fatalities, property damage and crop damage. Also interesting 
to note is that event types like heat seem hazardous to health but not so much for
property and crops. Conversely, things like hail and thunderstorm winds affect crop
and property but they affect heath to a lesser extent.

The harm value is a made up value to find the most relavant EVTYPEs. We don't need 
it anymore.
```{r}
evtypes <- evtypes[, !"harm"]
```

We want to know what percent of the total harm do these 10 account for
```{r}
toptypes <- head(evtypes, 10)
toptypes[,
  list(
    injuries = sum(injuries)/totals$injuries,
    fatalities = sum(fatalities)/totals$fatalities,
    property.damage = sum(property.damage)/totals$property.damage,
    crop.damage = sum(crop.damage)/totals$crop.damage
  )
]
```

So our subset accounts for a majority of the harm reported in the data (at least 
67%). So now we will mostly focus on those 10 EVTYPEs. We will add a row to the 
subset summing up the contributions from the other EVTYPEs.
```{r}
others <- evtypes[!EVTYPE %in% toptypes$EVTYPE,]
otherstotal <- others[,
  list(
    EVTYPE = "OTHER",
    frequency = sum(frequency),
    injuries = sum(injuries),
    fatalities = sum(fatalities),
    property.damage = sum(property.damage),
    crop.damage = sum(crop.damage)
  )
]
evtypes <- rbind(toptypes,otherstotal)
options(width = 120)
evtypes
```

Here is a plot of the above data separated by the type of harm
```{r}
par(mar=c(2,3,2,1))
layout(matrix(c(1,2,5,3,4,5),2,3,byrow=TRUE))
library(RColorBrewer)
cols <- brewer.pal(11,"Set3")

with(evtypes, barplot(injuries, main="Injuries", col=cols))
with(evtypes, barplot(fatalities, main="Fatalities", col=cols))
with(evtypes, barplot(property.damage, main="Property Damage ($)", col=cols))
with(evtypes, barplot(crop.damage, main="Crop Damage ($)", col=cols))
plot(1, 1, type="n", ann=FALSE, yaxt='n', xaxt='n')
legend("center", evtypes$EVTYPE, fill=cols, box.lwd=0)
```

Note that a very large factor of injuries are due to tornadoes and a very large
factor of crop damage is due to hail. Also this plot doesn't give us the best 
picture of what causes fatalities.

### Trends

We want to find out how things have changed over time for our top 10 EVTYPEs
```{r}
others <- data[!EVTYPE %in% toptypes$EVTYPE]
evtypesbyyear <- aggregate(
  cbind(INJURIES, FATALITIES, PROPDMG, CROPDMG) ~ EVTYPE + year, 
  data = rbind(
    data[EVTYPE %in% toptypes$EVTYPE],
    others[, EVTYPE := "OTHER"]
  ), 
  sum, 
  drop = FALSE
)
evtypesbyyear$EVTYPE <- factor(
  evtypesbyyear$EVTYPE,
  levels = evtypes$EVTYPE
)
```

Here's the plot
```{r, fig.width=6, fig.height=10}
library(cowplot)
p1 <- ggplot(
  evtypesbyyear, 
  aes(x=year, y=INJURIES, fill=EVTYPE)
) + 
  geom_area(position="stack") + 
  scale_fill_manual(values=cols) +
  theme_bw(base_size=10)
p2 <- ggplot(
  evtypesbyyear, 
  aes(x=year, y=FATALITIES, fill=EVTYPE)
) + 
  geom_area(position="stack") +
  scale_fill_manual(values=cols) +
  theme_bw(base_size=10)
p3 <- ggplot(
  evtypesbyyear, 
  aes(x=year, y=PROPDMG, fill=EVTYPE)
) + 
  geom_area(position="stack") +
  scale_fill_manual(values=cols) +
  theme_bw(base_size=10)
p4 <- ggplot(
  evtypesbyyear, 
  aes(x=year, y=CROPDMG, fill=EVTYPE)
) + 
  geom_area(position="stack") +
  scale_fill_manual(values=cols) +
  theme_bw(base_size=10)
plot_grid(
  p1 + theme(legend.position="none") + ylab(""),
  p2 + theme(legend.position="none") + ylab(""),
  p3 + theme(legend.position="none") + ylab(""),
  p4 + theme(legend.position="none") + ylab(""),
  get_legend(p1 + theme(legend.position="bottom")),
  labels = c(
    "     Injuries",
    "    Fatalities", 
    " Property Damage", 
    "   Crop Damage"
  ),
  nrow = 5,
  rel_heights = c(1,1,1,1,.3)
)
```

The above plot shows that other than at a few discrete times, injuries and fatalities 
due to severe weather are decreasing over time while property and crop damage seems
to be slowly increasing.

### Tornadoes

As seen before, tornadoes are the leading cause of injuries
```{r}
tornadoes <- data[EVTYPE == "TORNADO",]
injuriesbytornadoes <- signif(tornadoes[,sum(INJURIES)]*100/totals$injuries,4)
```

Tornadoes cause `r injuriesbytornadoes`% of all injuries

Here is a breakdown of injuries by the fujita scale. 
```{r}
tornadoes[
  order(F),
  list(
    frequency = .N,
    injuries = sum(INJURIES),
    injuries.per.tornado = sum(INJURIES)/.N,
    factor.of.total = sum(INJURIES)/evtypes[EVTYPE=="TORNADO",injuries]
  ),
  by = F
]
```

Unsurprisingly, milder tornadoes are more frequent and severe tornadoes cause more
injuries per tornado on avergae. Note that tornadoes with F-scale 3 cause the most
number of injuries in total (7324) accounting for more than 31% of all injuries
due to tornadoes.

Let's see where these tornado's are located
```{r}
par(mar=c(1,1,2.5,1))
with(
  tornadoes, 
  smoothScatter(
    -LONGITUDE, LATITUDE, 
    xlim=c(-12600,-6500), ylim=c(2300,5200),
    xaxt='n', yaxt='n',
    xlab='', ylab='',
    main="Locations of Tornadoes"
  )
)
```

You can clearly see the shape of the Continental USA from the plot above which shows
the locations of the tornadoes reported in the data. For some reason most of the
tornadoes are in the Eastern half of the country.

### Hail

Hail is the leading cause of crop damage
```{r}
hail <- data[EVTYPE == "HAIL",]
cropdmgbyhail <- signif(hail[,sum(CROPDMG)]*100/totals$crop.damage,4)
```

Hail causes `r cropdmgbyhail`% of all crop damage

Here is a breakdown of crop damage by hail size. 
```{r}
hail.sizes <- cut(
  hail$MAG, 
  breaks=c(25,37,50,75,88,100,125,150,175,200,250,275,300,400,450,500),
  labels = c(
    "Pea","Small marble","Penny","Nickel","Quarter",
    "Half dollar", "Ping Pong", "Golf ball", "Lime", "Tennis ball",
    "Baseball", "Large Apple", "Softball", "Grapefruit", "Computer CD"
  )
)
hail[,size := hail.sizes]
hail[
  order(-MAG),
  list(
    frequency = .N,
    crop.damage = sum(CROPDMG),
    damage.per.event = sum(CROPDMG)/.N,
    factor.of.total = sum(CROPDMG)/evtypes[EVTYPE=="HAIL",crop.damage]
  ),
  by = size
]
```

Golf ball sized hail causes the most amount of crop damage accounting for almost
a quarter of hail-related crop damage.
